National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Beliefs

### 2020-11-19 Manuscript.tex
include latex

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  english,
  man]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Beliefs},
  pdfauthor={Ben Highland1, Everett L. Worthington2, Don E Davis3, Chris G. Sibley4, \& Joseph Bulbulia5},
  pdflang={en-EN},
  pdfkeywords={Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

@@

\title{National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Beliefs}
\author{Ben Highland\textsuperscript{1}, Everett L. Worthington\textsuperscript{2}, Don E Davis\textsuperscript{3}, Chris G. Sibley\textsuperscript{4}, \& Joseph Bulbulia\textsuperscript{5}}
\date{}

@@

The relationship between spirituality and psychological well-being is a matter of enduring fascination (Ellsworth \& Ellsworth, 2010). Most previous attempts to quantify the relationship between these two domains find a positive association: spirituality predicts well-being (Ginsburg, Quirt, Ginsburg, \& MacKillop, 1995; Koenig, 2010; Koenig, Associate Professor of Psychology and Religious Studies Michael E McCullough, PhD, McCullough, Larson, \& Adjunct Professor of Psychiatry David B Larson, 2001; Russinova, Wewiorski, \& Cash, 2002; Smith, McCullough, \& Poll, 2003; Smith et al., 2003; Zaza, Sellick, \& Hillier, 2005). However, critics have raised three credible challenges.

First, previous research has frequently operationalised spirituality using measures of well-being. For example, The \emph{Daily Spiritual Experience Scale}, which is widely used in spiritual mental health research, includes items such as \enquote{I feel thankful for my blessings} and \enquote{I feel deep inner peace or harmony} to measure spirituality (Underwood \& Teresi, 2002). As Koenig observes, it is unsurprising that spirituality predict well-being when the same items are used to measure both constructs. Koenig aptly describes this approach as \enquote{tautological} (Koenig, 2008). Indeed, recent systematic-review of spirituality/mental health research found that nearly 45\% of previous studies employe tautological scales (Garssen et al., 2016). The first challenge, then, is to define measures of spirituality that are not well-being measures dressed up in other words.

A second challenge is to disentangle concepts of spirituality from concepts of religion. The term \enquote{spirituality} was first used to describe the religious practices of ascetics and monks, however, more recently, the term has taken on a broader family of meanings (Swinton, 2001). Spirituality and health research does not converge on a unique set of meanings from the wider universe of popular meanings. Although it is commonplace across most areas of psychological science for researchers to explore different operationalisations of focal concepts, the psychology of spirituality has mostly been conducted with cross-sectional North American samples among participants who affiliate with traditional religion (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky, Ratner, \& Chiu, 2005; Smith et al., 2003; Visser, Garssen, \& Vingerhoets, 2010; Yonker, Schnabelrauch, \& Dehaan, 2012). As such, the extent to which the psychology of spirituality and health has repeated or extended the psychology of religion and health remains unclear. Notably, researchers who have operationalised spirituality constructs as distinct from religion constructs have observed a negative association between spirituality and subjective well-being (King et al., 2013). Thus to infer how spirituality affects subjective well-being, it is important to disentangle the concepts of spirituality from concepts of religion, and to carefully investigate features of contructs in populations where these dimensions overlap, partially overlap, and diverge.

A third challenge is to investigate the relationship between spirituality and subjective well-being in nationally diverse samples over individual life-spans. Reliable inference requires observing change within people who differ dimensions of spirituality over time, and assing how their lives turn out (Garssen \& Visser, 2016). Previous investigations of spirituality and well-being, including high-quality reviews and meta-analyses, have relied on cross-sectional samples (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky et al., 2005; Smith et al., 2003; Visser et al., 2010; Yonker et al., 2012). However, To address the key questions in spirituality psychological health research requires longitudinal investigations. Unfortunately the relevant datasets are rare.

Here, we addresses the three challenges of previous in the following ways: First, to avoid a tautological elision of \enquote{spirituality} and \enquote{well-being} we adopt a non-affective measure of spirituality as a \enquote{believe in a Spirit or Life Force.} On the other side, to measure subjective well-being, we use a well-validated scale for Personal Well-being and a well-validated scale for Life-Satisfaction (see Method). Second, to disentangle spiritual beliefs from traditional religious beliefs and skeptical beliefs we adopt a four-level categorical indicator combining the binary indicator of belief in a Spirit or Life Force with the binary indicator \enquote{Do you believe in some form of a spirit or life force?} The levels of this four-level categorical indicator for spiritual, religious, and skeptical beliefs are as follows: (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who nevertheless believe in a Spirit or Life Force; (3) Those who believe in a God but do not believe in a Spirit or Life force; (4) Those who who believe in a both God and a Spirit or Life Force. Clearly such an approach cannot simultaneously address all interests in previous spirituality/well-being research. For example we do not combine our cognitive measure of belief with behavioral measures or attitudinal measures. However, this cost to scope is paid for with benefits to precision. The spiritual indicators measure cognitive states in which distinct varieties of skeptical beliefs, traditional religious beliefs, and non-traditional spirit beliefs are straightforwardly disentangled, and measured within people. Third, to assess whether spirit beliefs lead to growth, stability, or decline in psychological well-being, we leverage longitudinal responses from a nationally diverse probability sample year over an nine--year period with nine measurement points (2010 to 2018 New Zealand Attitudes and Values Study, NZAVS).

@@

The Time 10 (2018) wave of the NZAVS included 47,951 respondents. We analyzed data from participants who responded to our survey at least three four times between Time 2 (2010) and Time 10 (2018), resulting in a sample of \(N\) = 21,705 (\(N\) = 20,979 with complete responses).

@@

@@

We measured the socio-economic status of participants' immediate (small area) neighborhood using the 2013 New Zealand Deprivation Index, which uses aggregate census information about the residents of small neighborhood-type units to assign a decile-rank index from 1 (most affluent) to 10 (most impoverished) (Atkinson, Salmond, \& Crampton, 2014). The index is based on a Principal Components Analysis of the following nine variables (in weighted order): the proportion of adults who received a means-tested benefit, household income, proportion not owning own home, proportion single-parent families, proportion unemployed, proportion lacking qualifications, proportion household crowding, proportion no telephone access, and proportion no car access. Thus, the index reflects the average level of deprivation for small neighborhood-type units (or small community areas of about 80--90 people each) across the entire country. Our sample had a mean deprivation index of 4.80 (SD = 2.79). Deprivation was centered at its mean and standardized. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 1.

@@

Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05). We analyze data from participants who responded to the NZAVS survey at least three four times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of 21,705 (N = 20,979 with complete responses). To model growth in subjective well-being we use generalized linear mixed models. In our models, we account for the non-linear effects of time within participants by including the interaction of years (a continuous variable) with Beliefs (a four-level categorical indicator). The baseline for this indicator is Skepticism about a Spirit/Life Forces and Skepticism about a God. To handle the dependencies and heterogeneity introduced from the repeated measures, we include individual ID as an effect modelled as random. We centered and scaled the education, deprivation, and political conservativism. We centred age was centred at and put into decade-long intervals.
The model equation(s) are described in Equation 1.

@@

Because missing values can result in biased estimates (Blackwell, Honaker, \& King, 2017), we multiply imputed 10 datasets using the AmeliaII package in R(Honaker et al., 2011b).We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences. Details of this analysis can be found at: \url{https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing}.

@@

\begin{figure}
\includegraphics[width=6.4in]{Figs/USE_2_coefficientplots-1} \caption{Combination Coefficient Plots for Personal Well-being and Life Satisfaction Spirit Belief Longitudinal Models}\label{fig:unnamed-chunk-1}
\end{figure}

@@

\[

@@

\]

@@

We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict personal well-being (PWI) with Years (0-9), Beliefs (four categories), Age.10yrs.C, Deprivation.S, Edu.S, Employed, EthnicCats, Male, Partner, Pol.Orient.S and Urban, withId as random effects. The model included Id as random effects. The model's intercept, corresponding to Years = 0, Beliefs = Skeptic, Age = (sample mean), Deprivation.S = (sample mean), Edu.S =(sample mean), Employed = Not Employed, EthnicCats = Euro, Male = Not Mail, Partner = No Partner, Pol.Orient.S = (sample mean),, Urban = Not Urban and Id = 2, is at 4.94 (SE = 0.02, 95\% CI {[}4.89, 4.99{]}, p \textless{} .001) is at 6.63 (SE = 0.03, 95\% CI {[}6.57, 6.70{]}, p \textless{} .001). Within this model:

\begin{itemize}
\tightlist
\item
  The effect of Years is positive is not reliable (beta = 2.66e-03, SE = 3.58e-03, std. beta = 3.19e-03, p = 0.458). We infer that Skeptics personal well-being is not increasing in the Skeptic population
\item
  The effect of Beliefs in SpiritExcludesGod is reliably negative (beta = -0.14, SE = 0.03, std. beta = -0.03, p \textless{} .001). We infer that overall the population that believes in a Spirit but does not believe in a God is expected to have a lower personal well-being than skeptics.
\item
  The effect of Beliefs in GodAndSpirit is reliably negative (beta = -0.16, SE = 0.03, std. beta = -0.02, p \textless{} .001). We infer that overall the population that believes in God and a spirit or life force is expected to have a lower Personal Well-being than skeptics.
\item
  The effect of Beliefs in GodExcludesSpirit is not reliably different from the skeptic population (beta = 0.02, SE = 0.05, std. beta = 0.02, p = 0.669).
\item
  The effect of Age.10yrs.C is reliably positive (beta = 0.08, SE = 7.38e-03, std. beta = 0.07, p \textless{} .001). Older people have greater personal well-beingwellbeing.
\item
  The effect of Deprivation (S) is reliably negative (beta = -0.14, SE = 7.50e-03, std. beta = -0.08, p \textless{} .001). Deprived people have lower personal well-beingwellbeing.
\item
  The effect of Education (S) is reliably positive (beta = 0.12, SE = 8.74e-03, std. beta = 0.07, p \textless{} .001). Educated people greater personal well-beingwellbeing
\item
  The effect of Employed is reliably positive (beta = 0.09, SE = 0.01, std. beta = 0.05, p \textless{} .001). Employed people have greater personal well-beingwellbeing.
\item
  The effect of identifying as Maori is reliably negative (beta = -0.30, SE = 0.03, std. beta = -0.18, p \textless{} .001).
\item
  The effect of identifying as Pacific is reliably negative (beta = -0.29, SE = 0.05, std. beta = -0.17, p \textless{} .001).
\item
  The effect of identifying as Asian is not reliable (beta = -0.08, SE = 0.05, std. beta = -0.05, p = 0.089).
\item
  The effect of identifying as Male is reliably negative (beta = -0.11, SE = 0.02, std. beta = -0.06, p \textless{} .001).
\item
  The effect of having a Partner is reliably positive (beta = 0.61, SE = 0.02, std. beta = 0.37, p \textless{} .001).
\item
  The effect of Political Conservativism (S) is reliably positive (beta = 0.02, SE = 5.67e-03, std. beta = 0.01, p \textless{} .001).
\item
  The effect of living in an Urban area is reliably negative (beta = -0.02, SE = 0.01, std. beta = -0.01, p \textless{} .05).
\end{itemize}

@@

\begin{figure}
\includegraphics[width=6.4in]{Figs/USEPWI_expected-1} \caption{Combination Coefficient Plots for Personal Well-being and Life Satisfaction Spirit Belief Longitudinal Models}\label{fig:unnamed-chunk-2}
\end{figure}

Focusing on the interaction of time with belief-type on personal well-being we observe:

\begin{itemize}
\tightlist
\item
  The effect of years \(\times\) belief in a SpiritExcludesGod\_ is reliably positive (beta = 0.02, SE = 4.86e-03, std. beta = 0.02, p \textless{} .01). We infer that, unlike the skeptic population, people with beliefs in a spirit or life force who do not believe in a God tend to grow in their personal well-being over time.
\item
  The effect of years \(\times\) in GodAndSpirit is reliably positive (beta = 0.02, SE = 4.41e-03, std. beta = 0.03, p \textless{} .001). We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their personal well-being over time
\item
  The effect of Years \(\times\) belief in GodExcludesSpirit is not reliable(beta = 1.96e-03, SE = 8.29e-03, std. beta = 2.36e-03, p = 0.813). Thus, God believers who doubt in a Spirit or Life Force start maintain higher personal well-being than do other groups, but they are not expected to grow in personal well-being over time.
\end{itemize}

@@

We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict Life Satisfaction (LIFESAT) with Years (0-9), Beliefs (four categories), Age.10yrs.C, Deprivation (S), Education (S), Employed, EthnicCats, Male, Partner, Pol.Orient.S and Urban, with Id as random effects. Standardized parameters were obtained by fitting the model on a standardized version of the dataset. The model's total explanatory power is substantial (conditional R2 = 0.69). The model's intercept, corresponding to Years = 0, Beliefs = Skeptic, Age = (sample mean),, Deprivation.S = (sample mean), Education =(sample mean), Employed = Not Employed, EthnicCats = Euro, Male = Not Mail, Partner = No Partner, Pol.Orient.S = (sample mean), Urban = Not Urban and Id = 2, is at 4.94 (SE = 0.02, 95\% CI {[}4.89, 4.99{]}, p \textless{} .001). Within this model:

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  english,
  man]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Beliefs},
  pdfauthor={Ben Highland1, Everett L. Worthington2, Don E Davis3, Chris G. Sibley4, \& Joseph A. Bulbulia5},
  pdflang={en-EN},
  pdfkeywords={Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

% Overwrite redefinition of paragraph and subparagraph by the default LaTeX template
% See https://github.com/crsh/papaja/issues/292
\makeatletter
\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-1em}%
  {\normalfont\normalsize\bfseries\itshape\typesectitle}}

\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-\z@\relax}%
  {\normalfont\normalsize\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
\makeatother

@@

\title{National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Beliefs}
\author{Ben Highland\textsuperscript{1}, Everett L. Worthington\textsuperscript{2}, Don E Davis\textsuperscript{3}, Chris G. Sibley\textsuperscript{4}, \& Joseph A. Bulbulia\textsuperscript{5}}
\date{}

@@

The relationship between spirituality and psychological well-being is a matter of enduring fascination (Ellsworth \& Ellsworth, 2010). Most previous attempts to quantify the relationship between these two domains find a positive association: spirituality predicts well-being (Ginsburg, Quirt, Ginsburg, \& MacKillop, 1995; Koenig, 2010; Koenig, Associate Professor of Psychology and Religious Studies Michael E McCullough, PhD, McCullough, Larson, \& Adjunct Professor of Psychiatry David B Larson, 2001; Russinova, Wewiorski, \& Cash, 2002; Smith, McCullough, \& Poll, 2003; Zaza, Sellick, \& Hillier, 2005). However, critics have raised three credible challenges.

First, previous research has frequently operationalised spirituality using measures of well-being. For example, The \emph{Daily Spiritual Experience Scale}, which is widely used in spiritual mental health research, includes items such as \enquote{I feel thankful for my blessings} and \enquote{I feel deep inner peace or harmony} to measure spirituality (Underwood \& Teresi, 2002). As Koenig observes, it is unsurprising that spirituality predict well-being when the same items are used to measure both constructs. Koenig aptly describes this approach as \enquote{tautological} (Koenig, 2008). Indeed, recent systematic-review of spirituality/mental health research found that nearly 45\% of previous studies employ tautological scales (Garssen et al., 2016). The first challenge, then, is to define measures of spirituality that are not well-being measures dressed up in other words.

A second challenge is to disentangle concepts of spirituality from concepts of religion. The term \enquote{spirituality} was first used to describe the religious practices of ascetics and monks, however, more recently, the term has taken on a broader family of meanings (Swinton, 2001). Spirituality and health research does not converge on a unique set of meanings from the wider universe of popular meanings. Although it is commonplace across most areas of psychological science for researchers to explore different operationalisations of focal concepts, the psychology of spirituality has mostly been conducted with cross-sectional North American samples among participants who affiliate with traditional religion (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky, Ratner, \& Chiu, 2005; Smith et al., 2003; Visser, Garssen, \& Vingerhoets, 2010; Yonker, Schnabelrauch, \& Dehaan, 2012). As such, the extent to which the psychology of spirituality and health has repeated or extended the psychology of religion and health remains unclear. Notably, researchers who have operationalised spirituality constructs as distinct from religion constructs have observed a negative association between spirituality and subjective well-being (King et al., 2013). Thus to infer how spirituality affects subjective well-being, it is important to disentangle the concepts of spirituality from concepts of religion, and to carefully investigate features of constructs in populations where these dimensions overlap, partially overlap, and diverge.

A third challenge is to investigate the relationship between spirituality and subjective well-being in nationally diverse samples over individual life-spans. Reliable inference requires observing change within people who differ dimensions of spirituality over time, and assessing how their lives turn out (Garssen \& Visser, 2016). Previous investigations of spirituality and well-being, including high-quality reviews and meta-analyses, have relied on cross-sectional samples (Ano \& Vasconcelles, 2005; Hackney \& Sanders, 2003; Sawatzky et al., 2005; Smith et al., 2003; Visser et al., 2010; Yonker et al., 2012). However, To address the key questions in spirituality psychological health research requires longitudinal investigations. Unfortunately the relevant datasets are rare.

Here, we addresses the three challenges of previous in the following ways: First, to avoid a tautological elision of \enquote{spirituality} and \enquote{well-being} we adopt a non-affective measure of spirituality as a \enquote{believe in a Spirit or Life Force.} On the other side, to measure subjective well-being, we use a well-validated scale for personal well-being and a well-validated scale for Life-Satisfaction (see Method). Second, to disentangle spiritual beliefs from traditional religious beliefs and skeptical beliefs we adopt a four-level categorical indicator combining the binary indicator of belief in a Spirit or Life Force with the binary indicator \enquote{Do you believe in some form of a spirit or life force?} The levels of this four-level categorical indicator for spiritual, religious, and skeptical beliefs are as follows: (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who nevertheless believe in a Spirit or Life Force; (3) Those who believe in a God but do not believe in a Spirit or Life force; (4) Those who who believe in a both God and a Spirit or Life Force. Clearly such an approach cannot simultaneously address all interests in previous spirituality/well-being research. For example we do not combine our cognitive measure of belief with behavioral measures or attitudinal measures. However, this cost to scope is paid for with benefits to precision. The spiritual indicators measure cognitive states in which distinct varieties of skeptical beliefs, traditional religious beliefs, and non-traditional spirit beliefs are straightforwardly disentangled, and measured within people. Third, to assess whether spirit beliefs lead to growth, stability, or decline in psychological well-being, we leverage longitudinal responses from a nationally diverse probability sample year over an nine--year period with nine measurement points (2010 to 2018 New Zealand Attitudes and Values Study, NZAVS).

@@

The Time 10 (2018) wave of the NZAVS included 47,951 respondents. We analyzed data from participants who responded to our survey at least three times between Time 2 (2010) and Time 10 (2018), resulting in a sample of \(N\) = 21,705 (\(N\) = 20,979 with complete responses).

@@

We included standard demographic indicators used in New Zealand Attitudes and Values Research: \href{https://cdn.auckland.ac.nz/assets/psych/about/our-research/nzavs/NZAVSTechnicalDocuments/NZAVS-Technical-Documents-e11-Default-Statistical-Standards.pdf}{NZAVS modelling standards}

@@

We measured the socioeconomic status of participants' immediate (small area) neighborhood using the 2013 New Zealand Deprivation Index, which uses aggregate census information about the residents of small neighborhood-type units to assign a decile-rank index from 1 (most affluent) to 10 (most impoverished) (Atkinson, Salmond, \& Crampton, 2014). The index is based on a Principal Components Analysis of the following nine variables (in weighted order): the proportion of adults who received a means-tested benefit, household income, proportion not owning own home, proportion single-parent families, proportion unemployed, proportion lacking qualifications, proportion household crowding, proportion no telephone access, and proportion no car access. Thus, the index reflects the average level of deprivation for small neighborhood-type units (or small community areas of about 80--90 people each) across the entire country. Our sample had a mean deprivation index of 4.80 (SD = 2.79). Deprivation was centered at its mean and standardized. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 1.

@@

Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05). We analyze data from participants who responded to the NZAVS survey at least three times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of 21,705 (N = 20,979 with complete responses). To model growth in subjective well-being we use generalized linear mixed models. In our models, we account for the non-linear effects of time within participants by including the interaction of years (a continuous variable) with Beliefs (a four-level categorical indicator). The baseline for this indicator is Skepticism about a Spirit/Life Forces and Skepticism about a God. To handle the dependencies and heterogeneity introduced from the repeated measures, we include individual ID as an effect modelled as random. We centered and scaled the education, deprivation, and political conservativism. We centred age was centred at and put into decade-long intervals.
The model equation(s) are described in Equation 1.

@@

Because missing values can result in biased estimates (Blackwell, Honaker, \& King, 2017), we multiply imputed 10 datasets using the AmeliaII package in R(Honaker et al., 2011b).We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences. Additionally the results were robust to filtering of participants of N \textgreater3 to N = 9 waves. Additionally, the findings reported below were robust to modeling only the effects of years and beliefs, without adjusting for the standard NZAVS demographic covariates. The supplement describing these analysis is located at: \url{https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing}.

@@

\begin{figure}
\centering
\includegraphics{/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Figs/USE_2_coefficientplots-1.png}
\caption{\label{fig:unnamed-chunk-1}Combination Coefficient Plots for personal well-being and life satisfaction longitudinal growth models}
\end{figure}

@@

{[}

@@

{]}

@@

We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict personal well-being (PWI) with years (0-9), beliefs (four categories), Age.10yrs.C, deprivation (S), education (S), employed, ethnic categories, male, partner, political conservativism (S) and urban, with Id as a random effect. The model's intercept, corresponding to years = 0, beliefs = Skeptic, age = (sample mean), deprivation.S = (sample mean), education (S) =(sample mean), employed = not employed, ethnic Categories = European, male = not male, partner = no partner, political conservativism (S) = (sample mean), urban = not urban, is at 4.94 (SE = 0.02, 95\% CI {[}4.89, 4.99{]}, p \textless{} .001) is at 6.63 (SE = 0.03, 95\% CI {[}6.57, 6.70{]}, p \textless{} .001). Within this model:

\begin{itemize}
\tightlist
\item
  The effect of Years is positive is not reliable (beta = 2.66e-03, SE = 3.58e-03, std. beta = 3.19e-03, p = 0.458). We infer that personal well-being did not increase in the Skeptic population between 2010 and 2018.
\item
  The effect of Beliefs in \emph{Spirit Excludes God} is reliably negative (beta = -0.14, SE = 0.03, std. beta = -0.03, p \textless{} .001). We infer that in 2010, the population that believed in a Spirit but did not believe in a God had lower personal well-being than did Skeptics.
\item
  The effect of Beliefs in \emph{God And Spirit} is reliably negative (beta = -0.16, SE = 0.03, std. beta = -0.02, p \textless{} .001). We infer that in 2010, the population that believed in a God and a Spirit or Life Force had lower Personal Well-being than did Skeptics.
\item
  The effect of Beliefs in \emph{God Excludes Spirit} is not reliable (beta = 0.02, SE = 0.05, std. beta = 0.02, p = 0.669). We infer that in 2010, the population that believed in a God and doubted a Spirit or Life Force similar levels of personal well-being to Skeptics.
\item
  The effect of Age.10yrs.C is reliably positive (beta = 0.08, SE = 7.38e-03, std. beta = 0.07, p \textless{} .001). Older people tend to have greater personal well-being.
\item
  The effect of Deprivation (S) is reliably negative (beta = -0.14, SE = 7.50e-03, std. beta = -0.08, p \textless{} .001). Deprived people tend to have lower personal well-being.
\item
  The effect of Education (S) is reliably positive (beta = 0.12, SE = 8.74e-03, std. beta = 0.07, p \textless{} .001). Educated people tend to have greater personal well-being.
\item
  The effect of Employed is reliably positive (beta = 0.09, SE = 0.01, std. beta = 0.05, p \textless{} .001).
\item
  The effect of identifying as M=\{a\}ori is reliably negative (beta = -0.30, SE = 0.03, std. beta = -0.18, p \textless{} .001).
\item
  The effect of identifying as Pacific is reliably negative (beta = -0.29, SE = 0.05, std. beta = -0.17, p \textless{} .001).
\item
  The effect of identifying as Asian is not reliable (beta = -0.08, SE = 0.05, std. beta = -0.05, p = 0.089).
\item
  The effect of identifying as male is reliably negative (beta = -0.11, SE = 0.02, std. beta = -0.06, p \textless{} .001).
\item
  The effect of having a partner is reliably positive (beta = 0.61, SE = 0.02, std. beta = 0.37, p \textless{} .001).
\item
  The effect of political conservativism (S) is reliably positive (beta = 0.02, SE = 5.67e-03, std. beta = 0.01, p \textless{} .001).
\item
  The effect of living in an Urban area is reliably negative (beta = -0.02, SE = 0.01, std. beta = -0.01, p \textless{} .05). T
\end{itemize}

@@

\includegraphics{Figs/USEPWI_expected-1.png}
Focusing on the interaction of time with belief-type on personal well-being we observe:

\begin{itemize}
\tightlist
\item
  The effect of years \(\times\) belief in a \emph{Spirit Excludes God} is reliably positive (beta = 0.02, SE = 4.86e-03, std. beta = 0.02, p \textless{} .01). We infer that, unlike the skeptic population, people with beliefs in a spirit or life force who do not believe in a God tend to grow in their personal well-being over time.
\item
  The effect of years \(\times\) in \emph{God And Spirit} is reliably positive (beta = 0.02, SE = 4.41e-03, std. beta = 0.03, p \textless{} .001). We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their personal well-being over time
\item
  The effect of Years \(\times\) belief in \emph{God Excludes Spirit} is not reliable(beta = 1.96e-03, SE = 8.29e-03, std. beta = 2.36e-03, p = 0.813). Thus, God believers who doubt in a Spirit or Life Force start maintain higher personal well-being than do other groups, but they are not expected to grow in personal well-being over time.
\end{itemize}

@@

We fitted a linear mixed model (estimated using REML and nloptwrap optimizer) to predict personal well-being (PWI) with years (0-9), beliefs (four categories), Age.10yrs.C, deprivation (S), education (S), employed, ethnic categories, male, partner, political conservativism (S) and urban, with Id as a random effect. Standardized parameters were obtained by fitting the model on a standardized version of the dataset. The model's total explanatory power is substantial (conditional R2 = 0.69). The model's intercept, corresponding to years = 0, beliefs = Skeptic, age = (sample mean), deprivation.S = (sample mean), education (S) =(sample mean), employed = not employed, ethnic Categories = European, male = not male, partner = no partner, political conservativism (S) = (sample mean), urban = not urban is at 4.94 (SE = 0.02, 95\% CI {[}4.89, 4.99{]}, p \textless{} .001). Within this model:
